---
title: "ggplot2 Basics"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Preparing the data

```{r install packages}
install.packages("tidyverse")
install.packages("palmerpenguins")
```

Once a package is installed, you can load it by running the `library()`
function with the package name inside the parentheses:

```{r load packages}
library(tidyverse)
library(palmerpenguins)
```

## Getting started with ggplot2

```{r}
data("penguins")
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))
ggplot(data=penguins, mapping=aes(x=flipper_length_mm, y=body_mass_g)) + geom_point()
# the above two lines of code work the same
```

**ggplot(data = penguins)**: In ggplot2, you begin a plot with the ggplot() function. The ggplot() function creates a coordinate system that you can add layers to. The first argument of the ggplot() function is the dataset to use in the plot. In this case, it’s “penguins.”

**+**: Then, you add a “+” symbol to add a new layer to your plot. You complete your plot by adding one or more layers to ggplot().

**geom_point()**: Next, you choose a geom by adding a geom function. The geom_point() function uses points to create **scatterplots**, the geom_bar function uses bars to create bar charts, and so on. In this case, choose the geom_point function to create a scatter plot of points. The ggplot2 package comes with many different geom functions. You’ll learn more about geoms later in this course.

**(mapping = aes(x = flipper_length_mm, y = body_mass_g))**: Each geom function in ggplot2 takes a mapping argument. This defines how variables in your dataset are mapped to visual properties. The mapping argument is always paired with the aes() function. The x and y arguments of the aes() function specify which variables to map to the x-axis and the y-axis of the coordinate system. In this case, you want to map the variable “flipper_length_mm” to the x-axis, and the variable “body_mass_g” to the y-axis. 

## Enhancing visualization with aes (aesthetics)

```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species))
```

Then we will try playing with the three common aesthetics - color, shape and size. We can also change the alpha (transparency).

```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, shape=species))
```

```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species, shape=species))
```

```{r}
ggplot(data = penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species, shape=species, size=species))
```

```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species, alpha=species))
```

```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g),color="purple")
# note the color should be OUTSIDE of the aes function if set to a constant value
```

## Exploring geoms

The geom_smooth can show general trend of the data.
```{r}
ggplot(data=penguins) + geom_smooth(mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

We can also combine two types of geoms, like this:
```{r}
ggplot(data=penguins) + geom_smooth(mapping=aes(x=flipper_length_mm, y=body_mass_g)) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species))
```

We can fit multiple lines, like this:
```{r}
ggplot(data=penguins) + geom_smooth(mapping=aes(x=flipper_length_mm, y=body_mass_g, linetype=species))
```

Here are a few more geom types.

```{r}
ggplot(data=penguins) + geom_jitter(mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=cut))
```

The color parameter changes the color of border. We will need to use the fill parameter to change the color of bars.
```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=cut, color=cut))
```

```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=cut, fill=cut))
```

We can also use the fill param to build a stacked bar chart.
```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=cut, fill=clarity))
```

## Subsetting with Facet

To facet the plot with a single variable, use facet_wrap()
```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_wrap(~species)
```

```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=clarity, fill=cut)) + facet_wrap(~cut)
```

To facet the plot with two variables, use facet_grid()
```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_grid(sex~species)
```

## Exploring theme

Adjust the text orientation to make labels easier to read
```{r}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=clarity, fill=cut)) + facet_wrap(~cut) + theme(axis.text.x = element_text(angle = 90))
```

## Labels and annotations

Use labs() to add titles and labels:
```{r}
ggplot(data=penguins) + geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g, color=species)) + labs(title="Palmer Penguins: Flipper Length vs. Body Mass", subtitle="Sample of Three Penguin Species", caption = "Data collected by Dr. Kristen Gorman") + annotate("text", x=220, y=3500, label="The Gentoos are the largest", color="purple", fontface="bold", size=4.5, angle=25)
```

## Saving the plot

It can be done in RStudio export function, but here is how it can be done with code:
```{r}
ggsave("penguin_species.png") # saves the last plot generated
```